Analyzing DECREASE trials for extent of data fabrication [working title]

### 2017-05-28 submission/manuscript.Rmd
Residual commit

---
title: "Analyzing DECREASE trials for extent of data fabrication"
author: "Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document: default
csl: ../bibliography/ama.csl
bibliography: ../bibliography/library.bib
---

The effect of beta-blockers on perioperative mortality in non-cardiac surgery for patients with coronary artery disease (CAD) has been controversial [@Coleg5210] due to findings of research  misconduct in two related clinical trials  [@commissie2011;@commissie2012;@commissie2013]. Three meta-analyses that included the trials subject to research misconduct concluded that beta-blockers decrease perioperative mortality [@Devereaux313;@Angeli2010;@bouri2014] whereas a meta-analysis that excluded the suspect trials concluded that beta-blockers increase perioperative mortality [@bouri2014]. In these studies, perioperative mortality was defined as the deathrate of patients during the period of the surgical procedure, which typically includes admission, anaesthesia, surgery, and recovery.

<!-- Should we quote the (translated) main conclusion form the reports? -->
The trials subject to research misconduct were the Dutch DECREASE-I and DECREASE-IV trials [@commissie2011;@commissie2012;@commissie2013]. The committees that investigated the integrity of the DECREASE trials reported that data fabrication was likely but that the extent of the data fabrication remained unclear [@commissie2011;@commissie2012;@commissie2013]. Moreover, the latest guidelines still recommend the usage of beta-blockers in the perioperative period in certain cases [@esc2014;@Fleishere278]. Considering these consequences, we aim to estimate the extent of data fabrication in the DECREASE studies [@dunkelgrun2009;@poldermans1999] to further the debate on using beta-blockers in the perioperative period of CAD patients undergoing non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on the provenance of the raw data but did not investigate the extent to which the DECREASE trials deviated from comparable trials. The provenance is primarily concerned with the origins of the data, verifying things such as (but not limited to) the informed consent and whether data corresponded to patient files. The committee reports did not neglect statistical evaluation however: a statistical expert evaluated the applicability of forensic statistical methods [@commissie2013] to evaluate results of trials separately (i.e., DECREASE-I, DECREASE-IV). Nonetheless, comparing across trials is a method that has previously been used to monitor trial data quality or to test for potential data anomalies [@Buyse1999-jq]. Moreover, this method has previously proven to be effective in detecting data fabrication [@Knepper2016-la]. Evaluating the DECREASE trials to other published trials studying the effectiveness of beta-blockers with respect to perioperative mortality could prove informative of the potential extent of the fabrication in the DECREASE trials.

<!-- nieuw  revise-->
However, the effectiveness of perioperative beta-blockade is obfuscated by both trials afflicted by data fabrication and the type of beta-blocker used in the trial. Various kinds of beta-blockers could vary in their pharmalogical effectiveness with respect to perioperative mortality, and anomalies in the effectiveness of the DECREASE trials might be due to differences in the beta-blockers used [@klei2015] and not purely due to data fabrication (given not all data points can be considered fabricated at this point).

To statistically investigate the evidence of data fabrication in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. First, we reproduced the findings from a 2014 meta-analysis [@bouri2014] that contained sufficient information to estimate the deviation of the DECREASE trials from other published trials on beta-blockers. We also include type of beta-blocker to inspect whether this is predictive of the effect of beta-blockers on perioperative mortality. Second, we evaluated the probability that the DECREASE trials occurred assuming no data fabrication. Third, we reversed this assumption and assumed that data fabrication did occur and estimated how many data points would have to be fabricated to reproduce the results of the DECREASE trials, if the other published trials are regarded as estimating the true effect of beta-blockers on perioperative mortality in CAD patients undergoing non-cardiac surgery.

# Step 1: reproducing meta-analysis of Bouri et al. (2014)

## Methods

@@

## Results

@@

kable(table1, caption = 'The original- and reproduced meta-analytic results based on the data provided in the 2014 meta-analysis by Bouri et al.')
```

@@

# Step 2: evaluating the veracity of DECREASE studies

@@

## Method

@@

# Step 3: estimating the amount of fabricated data

@@

>>>>

---
title: "Analyzing DECREASE trials for extent of data fabrication [working title]"
author: "Chris HJ Hartgerink, Gerben ter Riet, Marleen Kemper"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
  html_document: default
  word_document: default
csl: ../bibliography/ama.csl
bibliography: ../bibliography/library.bib
---

The effect of beta-blockers on perioperative mortality in non-cardiac surgery has been controversial [@Coleg5210] due to findings of research  misconduct in two related clinical trials  [@poldermans1999;@dunkelgrun2009;@commissie2011;@commissie2012;@commissie2013]. Three meta-analyses that included the trials subject to research misconduct concluded that beta-blockers decrease perioperative mortality [@Devereaux313;@Angeli2010;@bouri2014] whereas a meta-analysis that excluded the suspect trials concluded that beta-blockers increase perioperative mortality [@bouri2014]. In these studies, perioperative mortality was defined as the deathrate of patients during the period of the surgical procedure, which typically includes admission, anaesthesia, surgery, and recovery.

The trials subject to research misconduct were the Dutch DECREASE-I and DECREASE-IV trials [@commissie2011;@commissie2012;@commissie2013]. The committees that investigated the integrity of the DECREASE trials reported that data fabrication was likely but that the extent of the data fabrication remained unclear [@commissie2011;@commissie2012;@commissie2013]. Moreover, the latest guidelines still recommend the usage of beta-blockers in the perioperative period in certain cases [@esc2014;@Fleishere278]. Considering these consequences, we aim to estimate the extent of data fabrication in the DECREASE studies [@poldermans1999;@dunkelgrun2009;@commissie2011;@commissie2012;@commissie2013] to further the debate on using beta-blockers in the perioperative period of CAD patients undergoing non-cardiac surgery.

The reports on the integrity of the DECREASE trials primarily focused on the provenance of the raw data but did not investigate the extent to which the DECREASE trials deviated from comparable trials. The provenance is primarily concerned with the origins of the data, verifying things such as (but not limited to) the informed consent and whether data corresponded to patient files. The committee reports did not neglect statistical evaluation however: a statistical expert evaluated the applicability of forensic statistical methods [@commissie2013] to evaluate results of trials separately (i.e., DECREASE-I, DECREASE-IV). Nonetheless, comparing across trials is a method that has previously been used to monitor trial data quality or to test for potential data anomalies [@Buyse1999-jq]. Moreover, this method has previously proven to be effective in detecting data fabrication [@Knepper2016-la]. Comparing the DECREASE trials to other published trials studying the effectiveness of beta-blockers with respect to perioperative mortality could prove informative of the potential extent of the fabrication in the DECREASE trials.

The effectiveness of perioperative beta-blockade is obfuscated by the afflicted DECREASE trials, potentially interacting with the type of beta-blocker and the way that beta-blocker was administered (i.e., dose and duration of treatment).  Throughout the clinical trials regarding beta-blockers, patients were administered various beta-blockers (e.g., metroprolol, bisoprolol, atenolol) and in various ways (e.g., intraveneously, orally; half an hour before surgery or multiple days before surgery). Factors such as dosage and duration can have an effect on the pharmacological effectiveness with respect to perioperative mortality. Moreover, the anomalous results from the DECREASE trials might partly be caused by such differences [@klei2015] and not purely due to data fabrication (given not all data points can be considered fabricated at this point).

The latest guidelines of the European Society of Cardiology and the American Heart Association still recommend the usage of beta-blockers in certain cases [@;@]. Therefore, we evaluate the usage of beta-blockers with respect to its effectiveness for the the perioperative period in patients undergoing non-cardiac surgery. First, we estimated the extent of data fabrication in the DECREASE studies [@poldermans1999;@dunkelgrun2009] Secondly, we evaluated the clinical studies used in the different meta-analysis from a pharmacological point of view.

# Part 1: statistical estimation of data fabrication in DECREASE trials

To statistically investigate the evidence of data fabrication in the DECREASE studies [@dunkelgrun2009;@poldermans1999], we took three steps. First, we reproduced the findings from the 2014  meta-analysis by Bouri et al. [@bouri2014] that contained sufficient information to estimate the deviation of the DECREASE trials from other published trials on beta-blockers. We also include type of beta-blocker to inspect whether this is predictive of the effect of beta-blockers on perioperative mortality. Second, we evaluated the probability that the DECREASE trials occurred assuming no data fabrication. Third, we reversed this assumption and assumed that data fabrication did occur and estimated how many data points would have to be fabricated to reproduce the results of the DECREASE trials, if the other published trials are regarded as estimating the true effect of beta-blockers on perioperative mortality in patients undergoing non-cardiac surgery.

## Step 1: reproducing meta-analysis of Bouri et al. (2014)

### Methods

@@

### Results

@@

knitr::kable(table1, caption = '__Table XX.__ The original- and reproduced meta-analytic results based on the data provided in the 2014 meta-analysis by Bouri et al.')
```

@@

## Step 2: evaluating the veracity of DECREASE studies

@@

### Method

@@

## Step 3: estimating the amount of fabricated data

@@

# Part 2:
